---
title: "ArmedConflictEDA"
author: "Xinze Yu"
format: pdf
editor: visual
---

## Study Objectives and Operationalizations

The **objective** of the current analysis is to study how armed conflicts impact the maternal and child health from a global scope.\

Specifically, the primary **exposure variable** of interest is **armed conflict**, as defined by the UCDP, it is a a binary variable indicating the presence of conflict for each country--year observation (0 = no, $<25$ battle-related deaths; 1 = yes, $\le25$ battle-related deaths).\

The primary **outcome** measures are maternal, under-5, infant, and neonatal **mortality rates**.\

A list of covariates is included in the dataset, and will potentially be included in the model: "gdp1000", "OECD", "OECD2023", "popdens", "urban", "agedep", "male_edu", "temp", "rainfall1000", "Drought", "Earthquake".\

## 1. Explore Data Structure and Summary Statistics of Key Variables

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load required packages
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(RColorBrewer)
```

We can start from checking the overall structure, data type, and missing values.

```{r}
# import data
acdata <- read.csv(here('data', 'analytical', 'finaldata.csv'), header = TRUE)
# factorize primary binary exposure
acdata$armed_conflict <- as.factor(acdata$armed_conflict)

# check structure, summary statistics, and missingness
str(acdata)
summary(acdata)
```
Note that one of the key outcome variables, maternal mortality, has noticeable missingness - 426 (11.5%) observations were missing out of 3720 observations.\

## 2. Visualize the Distributions

Plot **histograms** for outcome variables to identify skewness or outliers.

```{r, warning=FALSE}
# define outcome
outcomes <- c("MatMortality", "Und5Mortality", "InfMortality", "NeoMortality")

# create empty list to store plots
plot_list <- list()

# create histogram for each outcome
for (var in outcomes) {
  p <- ggplot(acdata, aes_string(x = var)) + 
    geom_histogram(binwidth = 30, fill = "blue", color = "black") + 
    labs(title = paste("Distribution of", var), x = var, y = "Count") + 
    theme(plot.title = element_text(size = 10)) 
  plot_list[[var]] <- p
}

# arrange plots into grid layout
grid.arrange(grobs = plot_list, ncol = 2)
```

The largely empty graph in maternal mortality suggests the presence of outliers. We can proceed to use logarithms.

```{r, warning=FALSE}
# create histogram for maternal mortality with log scale
ggplot(acdata, aes_string(x = "MatMortality")) + 
    geom_histogram(binwidth = 30, fill = "blue", color = "black") + 
    scale_y_log10() + 
    labs(title = paste("Distribution of", "MatMortality", "(with a log scale)"), 
         x = "MatMortality", y = "Count")
```

```{r, warning=FALSE}
# create empty list to store plots
plot_list1 <- list()

# create boxplot to compare each outcome between 
# conflict and non-conflict countries
for (var in outcomes) {
  p <- ggplot(acdata, aes_string(x = "armed_conflict", y = var)) + 
    geom_boxplot() + 
    labs(title = paste(var, "by Conflict Status"), 
         x = "Armed Conflict (0=No, 1=Yes)", y = var)
  plot_list1[[var]] <- p
}

# arrange plots into grid layout
grid.arrange(grobs = plot_list1, ncol = 2)
```
Overall, the country/year with armed conflicts shows higher mortality rates among all outcome groups. 

## 3. Explore Relationships between Variables

```{r}
# define covariates
covariates <- c("gdp1000", "OECD", "OECD2023", "popdens", 
                "urban", "agedep", "male_edu", "temp", 
                "rainfall1000", "Drought", "Earthquake")

# correlation between key variable
cor.mat <- cor(acdata[, c(outcomes, covariates)], use = "complete.obs")

# melt the correlation matrix to long format for ggplot
melted.cor.mat <- melt(cor.mat)

# Draw the heatmap using ggplot
ggplot(data = melted.cor.mat, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile(color = "white") + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  coord_fixed() + 
  labs(title = "Heatmap of Correlations Between Outcomes and Covariates", 
       x = "", y = "")
```

## Summary of the EDA on Armed Conflict and Maternal & Child Health

**1. Data Structure and Summary**\

The dataset contains 3,720 observations across 21 variables, representing multiple countries over the period 2000–2019.
Key variables include maternal mortality, infant mortality, and under-5 mortality, which show insignificant levels of missingness, with maternal mortality having highest rate of missingness (11.5%) that might be concerning.\

**2. Distribution of Key Variables**\

Histograms of the outcome variables reveal highly skewed distributions with potential outliers for maternal mortality. To manage this skewness, a log scale transformation was applied to the maternal mortality data, which provided a clearer view of the distribution.\

Boxplots comparing conflict vs. non-conflict countries indicate higher mortality rates for maternal, under-5, infant, and neonatal mortality in conflict-affected countries. The presence of conflict appears to show a noticable association with worsened health outcomes for mothers and children.\

**3. Correlations and Relationships**\

A correlation heatmap was created to examine the relationships between the outcome variables and covariates. Key findings include: strong negative correlations between male_edu and all mortality measures; strong positive correlations between all mortality rates and agedep and temp. No significant correlations between natural disasters (e.g., drought, earthquakes) and the mortality outcomes.\



