---
title: "ArmedConflictEDA"
author: "Xinze Yu"
format: pdf
editor: visual
---

## Study Objectives and Operationalizations

The **objective** of the current analysis is to study how armed conflicts impact the maternal and child health from a global scope. \

Specifically, the primary **exposure variable** of interest is **armed conflict**, as defined by the UCDP, it is a a binary variable indicating the presence of conflict for each country–year observation (0 = no, $<25$ battle-related deaths; 1 = yes, $\le25$ battle-related deaths).\

The primary **outcome** measures are maternal, under-5, infant, and neonatal **mortality rates**.\

A list of covariates is included in the dataset, and will potentially be included in the model: "gdp1000", "OECD", "OECD2023", "popdens", "urban", "agedep", "male_edu", "temp", "rainfall1000", "Drought", "Earthquake".\


## 1. Explore Data Structure and Summary Statistics of Key Variables

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load required packages
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
```

We can start from checking the overall structure, data type, and missing values. 
```{r}
# import data
acdata <- read.csv(here('data', 'analytical', 'finaldata.csv'), header = TRUE)

# check structure, summary statistics, and missingness
str(acdata)
summary(acdata)
```
Note that one of the key outcome variables, maternal mortality, has noticeable missingness - 426 (11.5%) observations were missing out of 3720 observations. \

## 2. Visualize the Distributions

Plot **histograms** for outcome variables and covariates to identify skewness or outliers.
```{r, warning=FALSE}
# define outcome
outcomes <- c("MatMortality", "Und5Mortality", "InfMortality", "NeoMortality")

# create empty list to store plots
plot_list <- list()

# create histogram for each outcome
for (var in outcomes) {
  p <- ggplot(acdata, aes_string(x = var)) + 
    geom_histogram(binwidth = 30, fill = "blue", color = "black") + 
    labs(title = paste("Distribution of", var), x = var, y = "Count") + 
    theme(plot.title = element_text(size = 10)) 
  plot_list[[var]] <- p
}

# arrange plots into grid layout
grid.arrange(grobs = plot_list, ncol = 2)
```
The largely empty graph in maternal mortality suggests the presence of outliers. We can proceed to use logarithms. 

```{r, warning=FALSE}
# create histogram for maternal mortality with log scale
ggplot(acdata, aes_string(x = "MatMortality")) + 
    geom_histogram(binwidth = 30, fill = "blue", color = "black") + 
    scale_y_log10() + 
    labs(title = paste("Distribution of", "MatMortality", "(with a log scale)"), 
         x = "MatMortality", y = "Count")
```


To be finished...






